The Inconsistency in G\"odel's Ontological Argument: \\ A Success Story for AI in Metaphysics

### 2016-01-29 Papers/InProgress/IJCAI/submission.tex
minor fixes

Another novel contribution reported here (in
\S\ref{sec:improvedembedding}) is the implementation of an alternative
embedding for the slightly stronger modal logic \SFiveU, which is
based on a universal accessibility relation. Our experiments have
shown that the new embedding is more efficient, as the following two,
previously open problems can now be solved:
\begin{itemize}
\item Proving automatically the final theorem T3 \textit{Necessarily, there
    exists God} (in Scott's \shortcite{ScottNotes} consistent variant) directly from the axioms
  alone, without relying on the argument's intermediate argumentation
  steps (i.e., lemmata). \marginpar{ToDo: 2004 is the year of Sobel's book, not of Scott's manuscript. We need to fix this as I fixed the date for Goedel's manuscript. But I don't know the correct date for Scott's manuscript.}
\item Verifying automatically (in Isabelle/HOL) the proof of the modal
  collapse \cite{Sobel}, which is one of the most strongly criticized
  `side-effects' of the argument's axioms.
\end{itemize}

@@

For more than four decades, the serious consequences of G\"odel's
mistake remained unnoticed, despite intense activity on
criticizing his argument. Especially since the discovery by Sobel
\shortcite{Sobel} that modal collapse (MC)\footnote{The modal collapse,
  $\phi\rightarrow \Box \phi$, states that contingent truth implies
  necessary truth; it can be interpreted as \textit{everything is
    pre-determined} or even \textit{there is no free will}. } is
entailed by G\"odel's (or also Scott's) axioms, several variants have
been proposed
\cite{Anderson,AndersonGettings,Hajek1,Hajek2,Hajek3,Bjordal}
attempting to avoid the modal collapse. Many of these variants omit
the crucial conjunct in the definition of essence as well.\footnote{As
  these variants also change other axioms, on which the inconsistency
  of G\"odel's axioms depends, it is not necessarily the case that
  these variants are also inconsistent; they to be analyzed
  separately.} Opponents of the argument
(e.g. Oppy [\citeyear[p.226--227]{oppy96:_goedel_ontol_argum};
\citeyear[p.364]{oppy00:_respon_gettin};
\citeyear[p.1068]{oppy08:_higher_order_ontol_argum}]) have also proposed
parodies and other criticisms, referring to variants where the
conjunct is omitted.

@@

\subsection{Argument Reconstruction in Isabelle}  \label{sec:arg2}
To verify the correctness of the informal argument explained above, it
was reconstructed in Isabelle/HOL, using Metis\footnote{Metis, unlike
  external provers such as \textsc{Leo-II} or Satallax, 
  constructs proofs in Isabelle's highly trusted kernel calculus.} to automate the
inessential parts (cf. Fig.~\ref {InconsistencyIsabelleK}). The essential use of the Empty Essence Lemma, on
the other hand, is explicitly stated, to ensure that Isabelle is
reconstructing the same argument. In fact, without the help of this
lemma, Metis is still not strong enough to refute G\"odel's
axioms.
\begin{figure}[t]
\centerline{\includegraphics[width=1\columnwidth]{./Images/InconsistencyIsabelleK.png}}
\caption{Inconsistency of G\"odel's Axioms in HOML \K verified in Isabelle/HOL} \label{InconsistencyIsabelleK}
\end{figure}

@@

\subsection{Inconsistency of G\"odel's Axioms in \SFiveU}

Isabelle/HOL's Sledgehammer tool, which orchestrates calls to
external provers such as \textsc{Leo-II}, still
fails to detect the inconsistency of G\"odel's axioms (in the standard
embedding), while a direct modeling of the problem in TPTP THF syntax
in combination with a direct call of \textsc{Leo-II} succeeded. In
other words, without independent experiments directly in TPTP THF the
inconsistency would thus not have been detected.
\begin{figure*}[t]
  \centering
  \begin{subfigure}[t]{0.715\textwidth}
    \includegraphics[width=\textwidth]{./Images/Manuscript2.png}
    \caption{G\"{o}del's manuscript, with mutually inconsistent axioms and
      definitions highlighted} \label{GoedelScript} 
  \end{subfigure}
   \begin{subfigure}[t]{0.28\textwidth}
     \includegraphics[width=\textwidth]{./Images/Inconsistency_S5U_direct.png}
     \caption{Inconsistency in HOML \SFiveU} \label{Inconsistency_S5U} 
   \end{subfigure}
% \includegraphics[width=.73\textwidth]{./Images/Manuscript2.png} \hfill
% \includegraphics[width=.26\textwidth]{./Images/Inconsistency_S5U_direct.png}
% \caption{(a) G\"{o}del's manuscript, with mutually inconsistent axioms and
%   definitions highlighted; (b) Inconsistency in \SFiveU} \label{GoedelScript} \label{Inconsistency_S5U} 
\end{figure*}

>>>>

Another novel contribution reported here (in
\S\ref{sec:improvedembedding}) is the implementation of an alternative
embedding for the slightly stronger modal logic \SFiveU, which is
based on a universal accessibility relation. Our experiments have
shown that the new embedding is more efficient, as the following two,
previously open problems can now be solved:
\begin{itemize}
\item Proving automatically the final theorem T3 \textit{Necessarily, there
    exists God} (in Scott's \shortcite{ScottNotes} consistent variant) directly from the axioms
  alone, without relying on the argument's intermediate argumentation
  steps (i.e., lemmata). 
% \marginpar{ToDo: 2004 is the year of Sobel's book, not of Scott's manuscript. We need to fix this as I fixed the date for Goedel's manuscript. But I don't know the correct date for Scott's manuscript.}
\item Verifying automatically (in Isabelle/HOL) the proof of the modal
  collapse \cite{Sobel}, which is one of the most strongly criticized
  `side-effects' of the argument's axioms.
\end{itemize}

@@

For more than four decades, the serious consequences of G\"odel's
mistake remained unnoticed, despite intense activity on
criticising his argument. Especially since the discovery by Sobel
\shortcite{Sobel} that modal collapse (MC)\footnote{The modal collapse,
  $\phi\rightarrow \Box \phi$, states that contingent truth implies
  necessary truth; it can be interpreted as \textit{everything is
    pre-determined} or even \textit{there is no free will}. } is
entailed by G\"odel's (or also Scott's) axioms, several variants have
been proposed
\cite{Anderson,AndersonGettings,Hajek1,Hajek2,Hajek3,Bjordal}
attempting to avoid the modal collapse. Many of these variants omit
the crucial conjunct in the definition of essence as well.\footnote{As
  these variants also change other axioms, on which the inconsistency
  of G\"odel's axioms depends, it is not necessarily the case that
  these variants are also inconsistent; they to be analyzed
  separately.} Opponents of the argument
(e.g. Oppy [\citeyear[p.226--227]{oppy96:_goedel_ontol_argum};
\citeyear[p.364]{oppy00:_respon_gettin};
\citeyear[p.1068]{oppy08:_higher_order_ontol_argum}]) have also proposed
parodies and other criticisms, referring to variants where the
conjunct is omitted.

@@

\subsection{Argument Reconstruction in Isabelle}  \label{sec:arg2}
\begin{figure}[t]
\centerline{\includegraphics[width=1\columnwidth]{./Images/InconsistencyIsabelleK.png}}
\caption{Inconsistency of G\"odel's Axioms in HOML \K verified in Isabelle/HOL} \label{InconsistencyIsabelleK}
\end{figure}
\begin{figure*}[t]
  \centering
  \begin{subfigure}[t]{0.715\textwidth}
    \includegraphics[width=\textwidth]{./Images/Manuscript2.png}
    \caption{G\"{o}del's manuscript, with mutually inconsistent axioms and
      definitions highlighted} \label{GoedelScript} 
  \end{subfigure}
   \begin{subfigure}[t]{0.28\textwidth}
     \includegraphics[width=\textwidth,height=7.2cm]{./Images/Inconsistency_S5U_direct.png}
     \caption{Inconsistency in HOML \SFiveU} \label{Inconsistency_S5U} 
   \end{subfigure}
% \includegraphics[width=.73\textwidth]{./Images/Manuscript2.png} \hfill
% \includegraphics[width=.26\textwidth]{./Images/Inconsistency_S5U_direct.png}
 \caption{The inconsistency in G\"{o}del's manuscript has been
   detected and verified by HOL ATPs} 
\end{figure*}
To verify the correctness of the informal argument explained above, it
was reconstructed in Isabelle/HOL, using Metis\footnote{Metis, unlike
  external provers such as \textsc{Leo-II} or Satallax, 
  constructs proofs in Isabelle's highly trusted kernel calculus.} to automate the
inessential parts (cf. Fig.~\ref {InconsistencyIsabelleK}). The essential use of the Empty Essence Lemma, on
the other hand, is explicitly stated, to ensure that Isabelle is
reconstructing the same argument. In fact, without the help of this
lemma, Metis is still not strong enough to refute G\"odel's
axioms.

@@

\subsection{Inconsistency of G\"odel's Axioms in \SFiveU}
Isabelle/HOL's Sledgehammer tool, which orchestrates calls to
external provers such as \textsc{Leo-II}, still
fails to detect the inconsistency of G\"odel's axioms (in the standard
embedding), while a direct modeling of the problem in TPTP THF syntax
in combination with a direct call of \textsc{Leo-II} succeeded. In
other words, without independent experiments directly in TPTP THF the
inconsistency would thus not have been detected.
